---
title: "Estimating Respondents with Doctoral Degrees Using the 2022 ACS Data"
author: "Group 38"
date: "2024-10-03"
format:
  pdf:
    toc: true
    toc-depth: 2
---

## **1. Introduction**

This document aims to analyze the 2022 American Community Survey (ACS) dataset from IPUMS. We will estimate the total number of respondents in each state who have a doctoral degree using Laplace’s ratio estimator. Finally, we'll compare these estimates to the actual number of respondents.

## **2. Instructions for Obtaining the Data**

To access the 2022 ACS data:

1. Visit the [IPUMS website]
2. Register for an account if you haven't already.
3. Once logged in, go to the [IPUMS USA](https://usa.ipums.org/usa/) section and select the 2022 ACS dataset.
4. Add the following variables to your extract:
    - `STATEICP`: State identifier based on IPUMS coding.
    - `EDUCD`: Educational attainment detail.
5. Download the `.csv.gz` file and decompress it using:
   ```bash
   gunzip usa_00002.csv.gz
   ```

## **3. Overview of the Ratio Estimator Approach**

The ratio estimator is a technique commonly used to estimate the size of a population based on known sample characteristics. Here, we use the following formula to estimate the total respondents in a state:

$$
\text{Estimated Total Respondents} = \frac{\text{Respondents with Doctoral Degrees}}{\text{Total Respondents in the State}} \times \text{Known Total in California}
$$


where:

- The numerator is the number of respondents with a doctoral degree in each state.

- The denominator is the total number of respondents in each state.

- The known total for California is provided as 391,171.

## **4. Data Analysis**

```{r}
# Load necessary libraries
library(dplyr)
library(readr)
library(here)
   
system("gunzip usa_00002.csv.gz")

# Read the dataset
acs_data <- read_csv(here("usa_00002.csv"))

# Filter for respondents with doctoral degrees
doctoral_data <- acs_data %>%
  filter(EDUCD == 116) %>%
  group_by(STATEICP) %>%
  summarize(doctoral_count = n())

# Count total respondents in each state
state_totals <- acs_data %>%
  group_by(STATEICP) %>%
  summarize(total_count = n())

# Join the two tables
state_data <- doctoral_data %>%
  left_join(state_totals, by = "STATEICP")

# Calculate the ratio between doctoral respondents and total respondents
state_data <- state_data %>%
  mutate(ratio = doctoral_count / total_count)

# Estimate the total number of respondents using the ratio and known total for California
california_total <- 391171
state_data <- state_data %>%
  mutate(estimated_total = ratio * california_total)

# Display the state data with estimates
print(state_data)
```

## **5. Comparison to Actual Respondent Totals**

```{r}
# Calculate the difference between actual and estimated totals
state_data <- state_data %>%
  mutate(difference = total_count - estimated_total)

# Display a summary of the differences
summary(state_data$difference)

# Visualize the comparison between actual and estimated counts
library(ggplot2)

ggplot(state_data, aes(x = total_count, y = estimated_total, label = STATEICP)) +
  geom_point(color = 'blue') +
  geom_text(nudge_x = 5000, nudge_y = 5000, size = 3) +
  labs(title = "Comparison of Actual vs. Estimated Respondent Totals by State",
       x = "Actual Total Respondents",
       y = "Estimated Total Respondents") +
  theme_minimal()
```

## **6. Observations and Explanation**

The discrepancies between the estimated and actual total respondents in each state can be attributed to various factors:

- **Sample Representation**: The proportion of respondents with doctoral degrees may not be consistent across states, leading to differences in estimates.

- **Variation in Ratios**: States may have varying educational demographics that aren't captured accurately when using a single ratio.

- **Size of the California Sample**: The estimation depends on the accuracy of the known respondent total in California, which may not generalize well across other states.

## **7. Conclusion**

The use of Laplace's ratio estimator provides a quick method to estimate population sizes based on sample characteristics. While useful, the results highlight that estimates can differ significantly from actual values, indicating the importance of considering state-specific demographics and context.

---

## **Appendix**

The dataset and code used in this analysis are available on [GitHub](https://github.com/DavidFJ207/2022_ACS_Ratio.git).

---