---
title: "Estimating Respondents with Doctoral Degrees Using the 2022 ACS Data"
author: "Gadiel Flores, Tina Kim, and Yanfei Huang"
date: "October 3, 2024"
abstract: This study aims to estimate the number of respondents with doctoral degrees across different states using the 2022 American Community Survey (ACS) dataset from IPUMS. We employ the Laplace ratio estimator, a statistical method that uses the proportion of respondents with doctoral degrees in California to estimate similar totals for other states. The analysis involves calculating the ratio of doctoral degree holders to total respondents for each state and applying this ratio to estimate state-specific totals. These estimates are then compared with the actual reported numbers. The findings underscore the importance of considering state-specific demographics and educational patterns, as discrepancies between estimated and actual totals highlight potential limitations of generalized estimation techniques. This work contributes to the broader understanding of educational attainment across the U.S. and illustrates the challenges in applying ratio estimators to diverse populations.
format:
  pdf:
    toc: true
    toc-depth: 2
bibliography: references.bib
---

## **1. Introduction**

This document aims to analyze the 2022 American Community Survey (ACS) dataset from IPUMS [@ipums]. This paper uses R [@citeR] for statistical computing. To efficiently manipulate and transform the data, we utilized the dplyr package [@dplyr]. For reading in and processing large CSV files, we used the readr package [@readr]. For managing file paths in our project, we used the here package [@here]. We will estimate the total number of respondents in each state who have a doctoral degree using Laplace’s ratio estimator [@laplace]. We applied a ratio estimator approach based on the methodology outlined in Cochran's book on sampling techniques [@cochran]. Finally, we compared these estimates to the actual number of respondents, finding that estimates may vary considerably from actual values, highlighting the need to account for state-specific demographics and contextual factors.

## **2. Instructions for Obtaining the Data**

To access the 2022 ACS data:

1. Visit the [IPUMS website]
2. Register for an account if you haven't already.
3. Once logged in, go to the [IPUMS USA](https://usa.ipums.org/usa/) section and select the 2022 ACS dataset.
4. Add the following variables to your extract:
    - `STATEICP`: State identifier based on IPUMS coding.
    - `EDUCD`: Educational attainment detail.
5. Download the `.csv.gz` file and decompress it using:
   ```bash
   gunzip usa_00002.csv.gz
   ```

## **3. Overview of the Ratio Estimator Approach**

The ratio estimator is a technique commonly used to estimate the size of a population based on known sample characteristics. Here, we use the following formula to estimate the total respondents in a state:

$$
\text{Estimated Total Respondents} = \frac{\text{Respondents with Doctoral Degrees}}{\text{Total Respondents in the State}} \times \text{Known Total in California}
$$


where:

- The numerator is the number of respondents with a doctoral degree in each state.

- The denominator is the total number of respondents in each state.

- The known total for California is provided as 391,171.

## **4. Data Analysis**

The dataset summarizes the distribution of respondents across 51 states, focusing on those with doctoral degrees (doctoral_count) and the total number of respondents (total_count). The ratio of doctoral degree holders to total respondents varies significantly between states. Using the known total of respondents in California as a baseline, the estimated_total column extrapolates the total respondents in each state.

```{r}
#| echo: false
#| message: false
#| warning: false

# Load necessary libraries
library(dplyr)
library(readr)
library(here)
   
system("gunzip usa_00002.csv.gz")

# Read the dataset
acs_data <- read_csv(here("data/usa_00002.csv"))

# Filter for respondents with doctoral degrees
doctoral_data <- acs_data %>%
  filter(EDUCD == 116) %>%
  group_by(STATEICP) %>%
  summarize(doctoral_count = n())

# Count total respondents in each state
state_totals <- acs_data %>%
  group_by(STATEICP) %>%
  summarize(total_count = n())

# Join the two tables
state_data <- doctoral_data %>%
  left_join(state_totals, by = "STATEICP")

# Calculate the ratio between doctoral respondents and total respondents
state_data <- state_data %>%
  mutate(ratio = doctoral_count / total_count)

# Estimate the total number of respondents using the ratio and known total for California
california_total <- 391171
state_data <- state_data %>%
  mutate(estimated_total = ratio * california_total)

# Display the state data with estimates
print(state_data)
```

## **5. Comparison to Actual Respondent Totals**

```{r}
#| echo: false
#| message: false
#| warning: false

#| fig-cap: "Actual vs. Estimated Respondent Totals by State."
#| fig-align: center

# Calculate the difference between actual and estimated totals
state_data <- state_data %>%
  mutate(difference = total_count - estimated_total)

# Display a summary of the differences
summary(state_data$difference)

# Visualize the comparison between actual and estimated counts
library(ggplot2)

ggplot(state_data, aes(x = total_count, y = estimated_total, label = STATEICP)) +
  geom_point(color = 'blue') +
  geom_text(nudge_x = 5000, nudge_y = 5000, size = 3) +
  labs(title = "Comparison of Actual vs. Estimated Respondent Totals by State",
       x = "Actual Total Respondents",
       y = "Estimated Total Respondents") +
  theme_minimal()
```

## **6. Observations and Explanation**

The discrepancies between the estimated and actual total respondents in each state can be attributed to various factors:

- **Sample Representation**: The proportion of respondents with doctoral degrees may not be consistent across states, leading to differences in estimates.

- **Variation in Ratios**: States may have varying educational demographics that aren't captured accurately when using a single ratio.

- **Size of the California Sample**: The estimation depends on the accuracy of the known respondent total in California, which may not generalize well across other states.

## **7. Conclusion**

The use of Laplace's ratio estimator provides a quick method to estimate population sizes based on sample characteristics. While useful, the results highlight that estimates can differ significantly from actual values, indicating the importance of considering state-specific demographics and context.

## **Appendix**

The dataset and code used in this analysis are available on [GitHub](https://github.com/DavidFJ207/2022_ACS_Ratio.git).

# References

